\documentclass{article}

\usepackage{url}
\usepackage{hyperref}

\title{Thesis Presentation Plan}
 
\author{Kyle McDonald}

\begin{document}

\maketitle

The first month of my composition ``Only Everything Lasts Forever'' will be publicly streamed online for a month. The stream itself will originate from a computer buried in a server room at EMPAC. The stream is being setup in conjunction with the EMPAC and RPI networking staff, with most technical-political hurdles already resolved (static IP address, server closet space, etc.). Anyone with a connection to the internet will be able to listen in on the composition for as long or short as they like.

The stream will begin on March 28th and end on April 28th. Some time during the first week of the stream there will also be an evening of discussion and refreshments in the West Hall Gallery (the sound composition will not be exhibited there). Information about the stream and the evening of discussion will be posted locally and online.

This configuration draws on my experiences sharing work online for the last six years, which has taken me from a centralized presentation of work on my website to a decentralized discussion happening across various websites, forums, and mailing lists. In 2003 I started maintaining a website with small generative code experiments, instruction pieces, and sound compositions. I received intermittent feedback via email, having discussions about code, process and aesthetics primarily with people I'd never met. Since then I've started posting to other places--Vimeo, Instructables, Flickr, various blogs and forums--and engaging in conversation through comment sections instead of via email. My website now contains only short summaries of projects and links to the decentralized, organically arising discussion that follows the sharing of documentation. Instead of serving as an exhibition space, the evening of discussion will act as a physical extension of the emergent ``virtual'' one.

A short description and link to the EMPAC hosted stream will appear on my website \url{http://kylemcdonald.net/}, while the development of the project and my written thesis is being documented on a Google Code repository. The repository can be monitored for changes at \url{http://code.google.com/p/oelf/source/list}.

\end{document}